A community-maintained standard library of population genetic models

### 2020-05-12 popSimManu.tex
Fixup table refs and editing pass.

For these reasons, we have generated a standardized, community-driven resource
for simulating published demographic models from a number of popular study systems.
This resource, which we call \stdpopsim, makes running
realistic simulations for population genetic analysis a simple matter of
choosing pre-implemented models from a community-maintained catalog.
The \stdpopsim catalog currently contains five species: humans,
\textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
\textit{Escherichia coli}, and \textit{Canis familiaris}.
For each species, the catalog contains information on our current understanding of
the physical organization of its genome (e.g., chromosome structure),
one or more inferred genetic maps (except for \textit{E.~coli}),
and default population-level parameters (mutation rate and generation time estimates).
The catalog also contains numerous published demographic models for the first three species.
These available models and parameters are meant to represent the field's current understanding,
and we intend for this resource to evolve as new results become available, and relevant
existing models are added to \stdpopsim by the community.
We have implemented both a command line interface and a simple Python API
that can be used to simulate genomic data
from a choice of organism, genetic map, chromosome, and demographic history.
In this way, \stdpopsim will lower the barrier to high-quality simulation for exploratory analyses,
enable rigorous evaluation of population genetic software,
and contribute to increased reliability of population genetic inferences.

The \stdpopsim library has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and maintain its quality.
Importantly, we developed rigorous quality control methods to ensure that we have
correctly implemented the models as described in their original publication
and provided documented methods for others to contribute new models.
We invite new collaborators to join our community:
those interested should visit our developer documentation at
\url{https://stdpopsim.readthedocs.io/en/latest/development.html}.
Below we describe the resource and give
examples of how it can be used to benchmark demographic inference methods.

@@

%ACS: the consortium really isn't the point; let's get right to describing what stdpopsim is
%The first contribution of the PopSim consortium is
The \stdpopsim library is a
a community-maintained collection of empirical genome data and population genetics simulation models,
with a structure depicted in Figure \ref{fig:cartoon}.
The package centers on a catalog of genomic information and demographic models
for a growing list of species (Fig.~\ref{fig:cartoon}A),
and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).

Given the genome data and simulation model descriptions defined within the
library, it is straightforward to run standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}.
The tree sequence format could also be converted
to other formats (e.g., VCF) by the user if desired.

@@

The central feature of \stdpopsim is the species catalog, a systematic organization
of the key quantitative data needed to simulate a given species.
Data is currently available for humans,
\textit{D.~melanogaster}, \textit{A.~thaliana},
\textit{Escherichia coli}, and \textit{Canis familiaris}.
A species definition consists of two key elements.
Firstly, the library defines some basic information about our current understanding of each
species' genome, including information about chromosome
lengths, average mutation rate estimates, and generation times.
We also provide access to detailed empirical information such as inferred genetic maps,
which model observed heterogeneity in recombination rate along chromosomes.
Such maps are often large,
so we do not distribute them directly with the software, but make them available
for download in a standard format.
When a simulation using such a map is requested by the user,
\stdpopsim will transparently download the map data into a local cache,
% (\texttt{$\sim$/.local/cache/stdpopsim} by default on Unix platforms)
where it can be quickly retrieved for subsequent simulations.
In the initial version of \stdpopsim we support
the HapMapII~\citep{international2007second} and
deCODE~\citep{kong2010fine} genetic maps for humans;
the \cite{salome2011recombination} map for \textit{A.~thaliana};
the \cite{comeron2012many} map for \textit{D.~melanogaster},
and the \cite{campbell2016pedigree} map for \textit{C.~familiaris}.
Adding further maps to the library is straightforward.
The second key element of a species description
within \stdpopsim is a set of carefully curated population genetic model
descriptions from the literature, which allow simulation under
specific historical scenarios that have been fit to present-day patterns of
genetic variation (See the Methods for a description of the community
development and quality-control process for these models.)

The current demographic models in the \stdpopsim catalog are shown in \autoref{tab:catalog}.
These range from
simple, single population histories \cite[e.g.,][]{sheehan2016deep},
to complex models which include population splitting, migration, and archaic
admixture \cite[e.g.,][]{ragsdale2019models}.
In addition to models from Table \autoref{tab:catalog},
the PopSim Consortium has demographic models for \textit{Pongo abelii}, \textit{Canis familiaris}, and \textit{Escherichia coli} in development at time of writing.

\renewcommand{\arraystretch}{1.2}
\begin{table}[t]
% Make sure the table is centered even though it's too wide
\makebox[\textwidth][c]{
    \begin{footnotesize}
    \input{model_table.tex}
    \end{footnotesize}
}
\caption{\label{tab:catalog}
Initial set of demographic models in the catalog and summary of computing resources needed for simulation.
For each model, we report the CPU time, maximum memory usage and the
size of the output \tskit file, as simulated using the \texttt{msprime} simulation engine.
In each case, we simulate 100 samples
drawn from the first population, for the shortest chromosome of that species
and a constant chromosome-specific recombination rate.
The times reported are for a single run on an Intel i5-7600K CPU.
Computing resources required will vary widely depending on sample sizes, chromosome length,
recombination rates and other factors.
}
\end{table}

Currently, {\em Homo sapiens} has the largest number of population models in
\stdpopsim (see Table \ref{tab:catalog}).
These models include: a simplified version of the \cite{tennessen2012evolution}
model with only the African population specified (expansion from the ancestral
population and recent growth; Africa\_1T12); the three-population model of \cite{gutenkunst2009inferring},
which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
the European and Asian populations (OutOfAfrica\_3G09); the \cite{tennessen2012evolution} two-population variant of the
\citeauthor{gutenkunst2009inferring} model, which does not include Asian populations but more explicitly models
recent rapid human population growth in Europe (OutOfAfrica\_2T12); the \cite{browning2018ancestry} admixture model
for American populations, which specifies ancestral African, European, and Asian population
components (AmericanAdmixture\_4B11); a three-population out-of-Africa model from \cite{ragsdale2019models},
which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19);
a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19);
and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).

@@

One of our goals is to engage research communities studying other taxa,
so as to expand the resource to many more species.
Although we have included demographic models and recombination maps,
there are many biological processes that we do not model.
Some of the additions that we are enthusiastic to add are:
selection (including distributions of fitness effects, maps of functional elements, and both single and recurrent hitchhiking events),
gene conversion, mutation models (rate heterogeneity),
more realistic demography (overlapping generations, separate sexes, mortality/fecundity schedules),
geographic population structure,
and downstream aspects of data quality (genotyping and mapping error).
Moreover, an in-depth investigation into the effects of population-size rescaling under many of the
above scenarios is warranted, given our preliminary findings using neutral simulations
(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
These plans would not be possible without recent important work on SLiM \citep{haller2019slim},
as well as current work on \texttt{msprime}.
Some other important processes are more challenging to model with current simulation software,
such as
structural variation,
changing recombination maps over time,
transposable elements,
and context-dependent mutation.

@@

\subsection*{Model quality control}
As a consortium we have agreed to a standardized procedure for model inclusion
into \stdpopsim that allows for rigorous quality control. Imagine Developer A
wants to introduce a new model into \stdpopsim. Developer A implements the
demographic model for the relevant organism along with clear documentation
of the model parameters and populations. This model is submitted as a ``pull request'',
where it is evaluated by a reviewer and then included as `preliminary',
but is not linked to the online documentation nor the command line interface.
Developer A submits a quality control (QC) issue, after which a second developer,
Developer B (perhaps found by requesting review from the broader Consortium),
then independently reimplements the model from the relevant
primary sources and adds an automatic unit test for equality between the
QC implementation and the preliminary production model.
If the two implementations are equivalent, the original model is included in \stdpopsim.
If not, we move to an arbitration process whereby A and B first try
to work out the details of what went wrong. If that fails, the original
authors of the published model must be contacted
to resolve ambiguities.
% This process can lead to cycles of revision of the implementation from Developers A and B but allows the project to assure quality.
% Indeed we have already had a number of interesting cases of arbitration to this point.
Further details of our QC process can be found in our developer documentation
(\url{https://stdpopsim.readthedocs.io/en/latest/development.html#}).

The possibility for error and the importance of careful qualty control
was inadvertantly illustrated during our own development process:
during the final revisions of this paper, we noticed that the
OutOfAfrica\_3G09 model \citep{gutenkunst2009inferring}
had not in fact gone through the QC process.
The subsequent QC process revealed that in fact our implementation
slightly wrong---migration rates had not been set to zero to the European population in the longest-ago time period
when there should have only been a single population.
This error was propagated from the \texttt{msprime} documentation,
where the model was presented as an illustrative example.
A number of studies have been published using copies of
this example code.

@@

\section*{Resource availability}
The version 0.1 release of \stdpopsim is available for download on Github:
\url{https://github.com/popsim-consortium/stdpopsim/releases}.
Documentation for the project can be found here:
\url{https://stdpopsim.readthedocs.io/en/latest/}.

>>>>

For these reasons, we have generated a standardized, community-driven resource
for simulating published demographic models from a number of popular study systems.
This resource, which we call \stdpopsim, makes running
realistic simulations for population genetic analysis a simple matter of
choosing pre-implemented models from a community-maintained catalog.
The \stdpopsim catalog currently contains six species: humans,
\textit{Pongo abelii},
\textit{Drosophila melanogaster}, \textit{Arabidopsis thaliana},
\textit{Escherichia coli}, and \textit{Canis familiaris}.
For each species, the catalog contains curated information on our current understanding of
the physical organization of its genome,
inferred genetic maps,
population-level parameters (e.g., mutation rate and generation time
estimates), and published demographic models.
These models and parameters are meant to represent the field's current understanding,
and we intend for this resource to evolve as new results become available, and
other existing models are added to \stdpopsim by the community.
We have implemented both a command line interface and a simple Python API
that can be used to simulate genomic data
from a choice of organism, genetic map, chromosome, and demographic history.
In this way, \stdpopsim will lower the barrier to high-quality simulation for exploratory analyses,
enable rigorous evaluation of population genetic software,
and contribute to increased reliability of population genetic inferences.

The \stdpopsim library has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and maintain quality.
Importantly, we developed rigorous quality control methods to ensure that we have
correctly implemented the models as described in their original publication
and provided documented methods for others to contribute new models.
We invite new collaborators to join our community:
those interested should visit our developer documentation at
\url{https://stdpopsim.readthedocs.io/en/latest/development.html}.
Below we describe the resource and give
examples of how it can be used to benchmark demographic inference methods.

@@

%ACS: the consortium really isn't the point; let's get right to describing what stdpopsim is
%The first contribution of the PopSim consortium is
The \stdpopsim library is
a community-maintained collection of empirical genome data and population genetics simulation models,
illustrated in Figure \ref{fig:cartoon}.
The package centers on a catalog of genomic information and demographic models
for a growing list of species (Fig.~\ref{fig:cartoon}A),
and software resources to facilitate efficient simulations (Fig.~\ref{fig:cartoon}B-C).

Given the genome data and simulation model descriptions defined within the
library, it is straightforward to run standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``succinct tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}.
The tree sequence format could also be converted
to other formats (e.g., VCF) by the user if desired.

@@

The central feature of \stdpopsim is the species catalog, a systematic organization
of the key quantitative data needed to simulate a given species.
Data is currently available for humans,
\textit{P.~abelii},
\textit{D.~melanogaster}, \textit{A.~thaliana},
\textit{E.~coli}, and \textit{C.~familiaris}.
A species definition consists of two key elements.
Firstly, the library defines some basic information about our current understanding of each
species' genome, including information about chromosome
lengths, average mutation rate estimates, and generation times.
We also provide access to detailed empirical information such as inferred genetic maps,
which model observed heterogeneity in recombination rate along chromosomes.
Such maps are often large,
so we do not distribute them directly with the software, but make them available
for download in a standard format.
When a simulation using such a map is requested by the user,
\stdpopsim will transparently download the map data into a local cache,
% (\texttt{$\sim$/.local/cache/stdpopsim} by default on Unix platforms)
where it can be quickly retrieved for subsequent simulations.
In the initial version of \stdpopsim we support
the HapMapII~\citep{international2007second} and
deCODE~\citep{kong2010fine} genetic maps for humans;
the \cite{nater2017morphometric} maps for \textit{P.~abelii};
the \cite{salome2011recombination} map for \textit{A.~thaliana};
the \cite{comeron2012many} map for \textit{D.~melanogaster},
and the \cite{campbell2016pedigree} map for \textit{C.~familiaris}.
Adding further maps to the library is straightforward.
The second key element of a species description
within \stdpopsim is a set of carefully curated population genetic model
descriptions from the literature, which allow simulation under
specific historical scenarios that have been fit to present-day patterns of
genetic variation (See the Methods for a description of the community
development and quality-control process for these models.)
\renewcommand{\arraystretch}{1.2}
\begin{table}[t]
% Make sure the table is centered even though it's too wide
\makebox[\textwidth][c]{
    \begin{footnotesize}
    \input{model_table.tex}
    \end{footnotesize}
}
\caption{\label{tab:catalog}
Initial set of demographic models in the catalog and summary of computing resources needed for simulation.
For each model, we report the CPU time, maximum memory usage and the
size of the output \tskit file, as simulated using the \texttt{msprime}
simulation engine (version 0.7.4).
In each case, we simulate 100 samples
drawn from the first population, for the shortest chromosome of that species
and a constant chromosome-specific recombination rate.
The times reported are for a single run on an Intel i5-7600K CPU.
Computing resources required will vary widely depending on sample sizes, chromosome length,
recombination rates and other factors.
}
\end{table}

The current demographic models in the \stdpopsim catalog are shown in \autoref{tab:catalog}.
% These range from
% simple, single population histories \cite[e.g.,][]{sheehan2016deep},
% to complex models which include population splitting, migration, and archaic
% admixture \cite[e.g.,][]{ragsdale2019models}.
{\em Homo sapiens} currently has the richest selection of population models.
These include: a simplified version of the \cite{tennessen2012evolution}
model with only the African population specified (expansion from the ancestral
population and recent growth; Africa\_1T12); the three-population model of \cite{gutenkunst2009inferring},
which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
the European and Asian populations (OutOfAfrica\_3G09); the \cite{tennessen2012evolution} two-population variant of the
\citeauthor{gutenkunst2009inferring} model, which does not include Asian populations but more explicitly models
recent rapid human population growth in Europe (OutOfAfrica\_2T12); the \cite{browning2018ancestry} admixture model
for American populations, which specifies ancestral African, European, and Asian population
components (AmericanAdmixture\_4B11); a three-population out-of-Africa model from \cite{ragsdale2019models},
which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19);
a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19);
and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).

@@

One of our goals is to engage research communities studying other taxa,
so as to expand the resource to many more species.
Although we have included demographic models and recombination maps,
there are many biological processes that we do not model.
Some of the additions that we are enthusiastic to add are:
selection (including distributions of fitness effects, maps of functional elements, and both single and recurrent hitchhiking events),
gene conversion, mutation models (rate heterogeneity),
more realistic demography (overlapping generations, separate sexes, mortality/fecundity schedules),
geographic population structure,
and downstream aspects of data quality (genotyping and mapping error).
Moreover, an in-depth investigation into the effects of population-size rescaling under many of the
above scenarios is warranted, given our preliminary findings using neutral simulations
(Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}).
Some other important processes are more challenging to model with current simulation software,
such as
structural variation,
changing recombination maps over time,
transposable elements,
and context-dependent mutation.

@@

\subsection*{Model quality control}
As a consortium we have agreed to a standardized procedure for model inclusion
into \stdpopsim that allows for rigorous quality control. Imagine Developer A
wants to introduce a new model into \stdpopsim. Developer A implements the
demographic model for the relevant organism along with clear documentation
of the model parameters and populations. This model is submitted as a ``pull request'',
where it is evaluated by a reviewer and then included as `preliminary',
but is not linked to the online documentation nor the command line interface.
Developer A submits a quality control (QC) issue, after which a second developer,
Developer B (perhaps found by requesting review from the broader Consortium),
then independently reimplements the model from the relevant
primary sources and adds an automatic unit test for equality between the
QC implementation and the preliminary production model.
If the two implementations are equivalent, the original model is included in \stdpopsim.
If not, we move to an arbitration process whereby A and B first try
to work out the details of what went wrong. If that fails, the original
authors of the published model must be contacted
to resolve ambiguities.
% This process can lead to cycles of revision of the implementation from Developers A and B but allows the project to assure quality.
% Indeed we have already had a number of interesting cases of arbitration to this point.
Further details of our QC process can be found in our developer documentation
(\url{https://stdpopsim.readthedocs.io/en/latest/development.html}).

The possibility for error and the importance of careful qualty control
was illustrated very clearly during our own development process:
while carrying out the final revisions of this paper, we noticed that the
OutOfAfrica\_3G09 model \citep{gutenkunst2009inferring}
had not gone through our QC process.
The subsequent QC revealed that our implementation was in fact
slightly wrong---migration rates had not been set to zero to the European population in the longest-ago time period
when there should have only been a single population.
This error was propagated from the \texttt{msprime} documentation,
where the model was presented as an illustrative example.
A number of studies have been published using copies of
this erroneous example code.

@@

\section*{Resource availability}
The \stdpopsim package is available for download on the Python Package Index:
\url{https://pypi.org/project/stdpopsim/}.
Documentation for the project can be found here:
\url{https://stdpopsim.readthedocs.io/en/latest/}.
